---
title: "all-sites-exploratory"
author: "Linda Herrera"
format: typst
execute:
  fig-format: png
---


## Packages

```{r}
#| label: packages
#| message: false
#| warning: false

library(tidyverse)
library(here)
library(janitor)
library(snakecase)
library(scales)
library(readxl)
library(patchwork)
library(ggridges)
library(lme4)
library(lmerTest)
library(knitr)
library(rmarkdown)

library(ragg)

```


## Data 

```{r}
#| label: cleaning and wrangling
#| message: false
#| warning: false

# Trail Creek Isotope 
compiled_isotope <- read_csv(here("data", "compiled_river_reach_isotope_data.csv"), show_col_types = FALSE) |> 
  clean_names() |> # lowercase and underscores in column names
  rename(sample = sample_name, # renaming columns to be shorter 
         delta_d = processed_delta_d,
         delta_d_std = processed_delta_d_stdev,
         delta_o = processed_delta_18o,
         delta_o_std = processed_delta_18o_stdev) |> 
  mutate(date = mdy(date)) # telling r the date format


gradient_data <- read_csv(here("data", "tc_gradient_data.csv"))

trail_creek <- compiled_isotope |> 
  filter(location %in% c("utc", "tc"))

coal_creek <- compiled_isotope |> 
  filter(location == "cc")

italian_creek <- compiled_isotope |> 
  filter(location == "ic")

```

## Exploratory Visuals with ALL Sites

Line-Conditioned Excess (lc-excess) vs Delta 18O 

```{r}
#| message: false
#| warning: false

ggplot(data = compiled_isotope, aes(x = delta_o, y = lc_excess, color = location)) +
  geom_point(size = 2.5, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.15, linewidth = 1) + 
  scale_color_manual(values = c("tc" = "turquoise3", "cc" = "sienna2", "ic" = "violetred2", "utc" = "blue")) +
  labs(
    title = "Line-Conditioned Excess (lc-excess) vs. Delta 18O",
    subtitle = "Comparing evaporation signals between sites",
    x = expression(paste(delta, " "^18, "O (\u2030)")), # Formats as δ18O (‰)
    y = "lc-excess (\u2030)",
    color = "Site Location") +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 14),
        legend.position = "right",
        panel.grid.minor = element_blank())

```

Time-Series of Line-Conditioned Excess (lc-excess)

```{r}
#| message: false
#| warning: false

ggplot(data = compiled_isotope, 
       aes(x = date, 
           y = lc_excess, 
           color = location, 
           group = location)) +
  geom_hline(yintercept = 0, 
             linetype = "dashed", 
             color = "gray50", 
             size = 0.8) +
  geom_point(size = 2.5, 
             alpha = 0.8) +
  geom_line(linewidth = 1, 
            alpha = 0.7) +
  labs(title = "Time-Series of Line-Conditioned Excess (lc-excess)",
       subtitle = "Values below 0 indicate potential evaporative enrichment",
       x = "Sampling Date",
       y = "lc-excess (\u2030)",
       color = "Site") +
  ylim(-8,5) +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 14),
        panel.grid.minor = element_blank(),
        legend.position = "top")

```

Time-Series of Deuterium Excess (d-excess)

```{r}
#| message: false
#| warning: false

ggplot(data = compiled_isotope, 
       aes(x = date, 
           y = d_excess, 
           color = location, 
           group = location)) +
  geom_hline(yintercept = 0, 
             linetype = "dashed", 
             color = "gray50", 
             size = 0.8) +
  geom_point(size = 2.5, 
             alpha = 0.8) +
  geom_line(size = 1, 
            alpha = 0.7) + 
  labs(title = "Time-Series of Deuterium Excess (d-excess)",
       subtitle = "Values below 0 indicate potential evaporative enrichment",
       x = "Sampling Date",
       y = "d-excess (\u2030)",
       color = "Site") +
  ylim(-3,15) +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 14),
        panel.grid.minor = element_blank(),
        legend.position = "top")

```

Temporal Variations in Line-Conditioned Excess (by location)

```{r}
#| label: line
#| message: false
#| warning: false

ggplot(compiled_isotope, aes(x = date, y = lc_excess, color = location)) +
  geom_point(size = 1.5) +
  labs(
    title = "Temporal Variations in Line-Conditioned Excess",
    x = "Date",
    y = "lc-excess (‰)",
    color = "Location"
  ) +
  facet_wrap(~location, scales = "free_y") +  # Compare across sites easily
  theme_minimal() + 
  theme(plot.title = element_text(face = "bold", size = 14))

```

Temporal Variations in Deuterium Excess (by sites)

```{r}
#| label: timeseries
#| message: false
#| warning: false

ggplot(compiled_isotope, aes(x = date, y = d_excess, color = location)) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1.5) +
  labs(
    title = "Temporal Variations in Deuterium Excess",
    x = "Date",
    y = "lc-excess (‰)",
    color = "Site"
  ) +
  facet_wrap(~site, scales = "free_y") +  # Compare across sites easily
  theme_minimal() + 
  theme(plot.title = element_text(face = "bold", size = 14))

```

Monthly Variations in Evaporative Fraction (lc-excess) Across Study Sites

```{r}
#| label: boxplot-monthly
#| message: false
#| warning: false

ggplot(compiled_isotope, aes(x = factor(month(date, label = TRUE, abbr = TRUE)), y = lc_excess)) +
  geom_boxplot(aes(fill = factor(month(date))), show.legend = FALSE) +
  facet_wrap(~location, scales = "free_y") +
 labs(
    title = "Monthly Variations in Evaporative Fraction (lc-excess) Across Study Sites",
    subtitle = "Seasonal isotopic shifts indicating changes in local evaporation-condensation histories",
    x = "Month",
    y = "lc-excess (‰)"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 11),
    axis.text.x = element_text(angle = 45, hjust = 1)) # tilting the x axis text

```

## Exploratory Visuals of Trail Creek

Line-Conditioned Excess (lc-excess) vs. Delta 18O

```{r}
#| label: scatterplot
#| message: false
#| warning: false

ggplot(data = trail_creek, aes(x = delta_o, y = lc_excess, color = site)) +
  geom_point(size = 2.5, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, alpha = 0.15, linewidth = 1) +
  labs(
    title = "Line-Conditioned Excess (lc-excess) vs. Delta 18O",
    subtitle = "Comparing evaporation signals between sites",
    x = expression(paste(delta, " "^18, "O (\u2030)")), # Formats as δ18O (‰)
    y = "lc-excess (\u2030)",
    color = "Site Location") +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 14),
        legend.position = "right",
        panel.grid.minor = element_blank())

```

Time-Series of Line-Conditioned Excess (lc-excess)

```{r}
#| label: time-series
#| message: false
#| warning: false

ggplot(data = trail_creek, 
       aes(x = date, 
           y = lc_excess)) +
  geom_hline(yintercept = 0, 
             linetype = "dashed", 
             color = "gray50", 
             size = 0.8) +
  geom_point(size = 2.5, 
             alpha = 0.8) +
  geom_line(size = 1, 
            alpha = 0.7) +
  labs(title = "Time-Series of Line-Conditioned Excess (lc-excess)",
       subtitle = "Values below 0 indicate potential evaporative enrichment",
       x = "Sampling Date",
       y = "lc-excess (\u2030)",
       color = "Site") +
  ylim(-8,5) +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 14),
        panel.grid.minor = element_blank(),
        legend.position = "top")

```

Time-Series of Deuterium Excess (d-excess)

```{r}
#| label: dexcess
#| message: false
#| warning: false

ggplot(data = trail_creek, 
       aes(x = date, 
           y = d_excess)) +
  geom_hline(yintercept = 0, 
             linetype = "dashed", 
             color = "gray50", 
             size = 0.8) +
  geom_point(size = 2.5, 
             alpha = 0.8) +
  geom_line(size = 1, 
            alpha = 0.7) + 
  labs(title = "Time-Series of Deuterium Excess (d-excess)",
       subtitle = "Values below 0 indicate potential evaporative enrichment",
       x = "Sampling Date",
       y = "d-excess (\u2030)",
       color = "Site") +
  ylim(-3,15) +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 14),
        panel.grid.minor = element_blank(),
        legend.position = "top")

```

Temporal Variations in Line-Conditioned Excess

```{r}
#| label: abc
#| message: false
#| warning: false

ggplot(trail_creek, aes(x = date, y = lc_excess, color = site)) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1.5) +
  labs(
    title = "Temporal Variations in Line-Conditioned Excess",
    x = "Date",
    y = "lc-excess (‰)",
    color = "Site"
  ) +
  facet_wrap(~site, scales = "free_y") +  # Compare across sites easily
  theme_minimal() + 
  theme(plot.title = element_text(face = "bold", size = 14))

```

Monthly Variations in Evaporative Fraction (lc-excess) Across Study Sites

```{r}
#| label: -monthly
#| message: false
#| warning: false

ggplot(trail_creek, aes(x = factor(month(date, label = TRUE, abbr = TRUE)), y = lc_excess)) +
  geom_boxplot(aes(fill = factor(month(date))), show.legend = FALSE) +
  facet_wrap(~location, scales = "free_y") +
 labs(
    title = "Monthly Variations in Evaporative Fraction (lc-excess) Across Study Sites",
    subtitle = "Seasonal isotopic shifts indicating changes in local evaporation-condensation histories",
    x = "Month",
    y = "lc-excess (‰)"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 11), # Emphasize site names
    axis.text.x = element_text(angle = 45, hjust = 1)   # Tilt month labels if they overlap
  )

```

Dual-isotope plot (δ18O vs δD) with LMWL reference

```{r}
#| label: dual
#| message: false
#| warning: false

# Trail Creek LMWL calculated from OPIC

lmwl_slope <- 7.94
lmwl_intercept <- 9.37 

ggplot(compiled_isotope, aes(x = delta_o, y = delta_d, color = site)) +
  geom_point(size = 2.5, alpha = 0.7) +
  geom_abline(slope = lmwl_slope, intercept = lmwl_intercept, 
              linetype = "dashed", color = "black", linewidth = 0.8) +
  facet_wrap(~location, labeller = labeller(location = c(tc = "Trail Creek", 
                                                           cc = "Coal Creek", 
                                                           ic = "Italian Creek"))) +
  labs(title = "Dual Isotope Plot: \u03b418O vs \u03b4D",
       subtitle = "Dashed line = Trail Creek LMWL",
       x = expression(paste(delta^18, "O (\u2030)")),
       y = expression(paste(delta, "D (\u2030)")),
       color = "Reach") +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 14),
        legend.position = "top")

```

Lc-excess boxplot by reach, faceted by watershed

```{r}
#| label: helo
#| message: false
#| warning: false

ggplot(compiled_isotope, aes(x = site, y = lc_excess, fill = site)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1.5) +
  facet_wrap(~location, labeller = labeller(location = c(tc = "Trail Creek", 
                                                           cc = "Coal Creek", 
                                                           ic = "Italian Creek"))) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(title = "lc-excess by Reach Position",
       subtitle = "Values below 0 indicate evaporative enrichment",
       x = "Reach", y = "lc-excess (\u2030)") +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 14),
        legend.position = "none")
```

Time series of lc-excess across season, colored by reach, faceted by watershed

```{r}
#| label: time
#| message: false
#| warning: false


ggplot(compiled_isotope, aes(x = date, y = lc_excess, color = site, group = interaction(site, location))) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(size = 2, alpha = 0.8) +
  geom_line(alpha = 0.6) +
  facet_wrap(~location, labeller = labeller(location = c(tc = "Trail Creek", 
                                                           cc = "Coal Creek", 
                                                           ic = "Italian Creek"))) +
  labs(title = "Seasonal Trends in lc-excess",
       subtitle = "2025 low-snowpack year",
       x = "Sampling Date", y = "lc-excess (\u2030)", color = "Reach") +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 14),
        legend.position = "top",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

Mixed model: lc-excess ~ site * location, with sample as random effect if repeated

```{r}
#| label: interaction plot
#| message: false
#| warning: false

# Mixed model
model_mixed <- lmer(lc_excess ~ site * location + (1 | date), data = compiled_isotope)
summary(model_mixed)
anova(model_mixed)  # Type III tests via lmerTest

ggplot(compiled_isotope, aes(x = location, y = lc_excess, color = site, group = site)) +
  stat_summary(fun = mean, geom = "point", size = 3) +
  stat_summary(fun = mean, geom = "line", linewidth = 1) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.1) +
  labs(title = "Interaction Plot: Reach \u00d7 Watershed",
       x = "Watershed", y = "Mean lc-excess (\u2030)", color = "Reach") +
  theme_bw()

```


```{r}

delta_summary <- compiled_isotope |> 
  group_by(site) |> 
  summarize(
    mean_lc = mean(lc_excess, na.rm = TRUE),
    se_lc = sd(lc_excess, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )


ggplot(delta_summary, aes(x = site, y = mean_lc, fill = site)) +
  geom_col(color = "black", width = 0.7) +
  geom_errorbar(
    aes(ymin = mean_lc - se_lc, ymax = mean_lc + se_lc), 
    width = 0.3, 
    color = "black"
  ) +
  theme_minimal() +
  labs(
    title = "Mean LC Excess by Site with Standard Error",
    x = "Sampling Site",
    y = "Mean LC Excess"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1), # Tilts labels so they don't overlap
    legend.position = "none"
  )




```



```{r}

# Build a lookup table of stream distances, then compute reach boundaries
distances <- tibble(
  site = c("TC-P3-US", "TC-P3-mid", "TC-P3-DS", "TC-BDA1",
           "TC-BDA1.5", "TC-BDA2", "TC-BDA3"),
  dist_m = c(57.6, 333, 654, 1160, 1620, 2240, 3150)
) %>%
  mutate(
    xmax = dist_m,
    xmin = lag(dist_m, default = 0)  # previous site's distance, 0 for first
  )


gradient_data_clean <- gradient_data %>%
  select(site, normalized_stream_distance_m,
         `Δlc-excess/Δx_6-24`, `Δlc-excess/Δx_6-30`) %>%
  pivot_longer(
    cols = c(`Δlc-excess/Δx_6-24`, `Δlc-excess/Δx_6-30`),
    names_to = "date",
    values_to = "gradient"
  ) %>%
  mutate(
    date = case_when(
      date == "Δlc-excess/Δx_6-24" ~ "2026-06-24",
      date == "Δlc-excess/Δx_6-30" ~ "2026-06-30"
    ),
    gradient = as.numeric(gradient) 
  ) %>%
  left_join(distances %>% select(site, xmin, xmax), by = "site") %>%
  mutate(
    site = factor(site, levels = distances$site)
  ) %>%
  filter(!is.na(gradient))


ggplot(gradient_data_clean) +
  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = 0, ymax = gradient),
            fill = "#4C72B0", color = "black", linewidth = 0.3) +
  geom_hline(yintercept = 0, color = "steelblue", linewidth = 0.4) +
  facet_wrap(~ date, nrow = 1) +
  scale_x_continuous(
    breaks = distances$dist_m,
    labels = distances$site
  ) +
  labs(
    x = "Normalized stream distance (m)",
    y = expression(Delta*"lc-excess/"*Delta*"x")
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

All reaches show lc-excess decreasing downstream, on both dates. Since lc-excess becoming more negative = more evaporative enrichment, this version says every single reach along Trail Creek is adding evaporative signal, with no recovery zones. That's a materially different story than "some reaches enrich, some reaches dilute" — it now reads as continuous, cumulative evaporative loss along the whole restored gradient.

Where the deepest enrichment is happening differs by date:

6-24: TC-P3-DS reach is the sharpest dip (~-2.2), essentially a local hotspot just downstream of the mid-P3 restoration point.
6-30: TC-BDA1.5→BDA2 is now the sharpest dip (~-2.15), while TC-P3-DS has relaxed to a much smaller magnitude (~-0.55) compared to 6-24.

TC-BDA3 is the outlier reach on both dates — it's close to flat (near-zero gradient), meaning that final reach adds almost no additional evaporative signal, regardless of date. That's consistent between both panels and is probably your most robust finding here: whatever's happening upstream, the lowest reach isn't contributing much additional fractionation.



## Exploratory Visuals of Coal Creek


## Exploratory Visuals of Italian Creek




