---
title: "all-sites-exploratory"
author: "Linda Herrera"
format: typst
execute:
  fig-format: png
---


## Packages

```{r}
#| label: packages
#| message: false
#| warning: false

library(tidyverse)
library(here)
library(janitor)
library(viridis)
library(wesanderson)
library(lme4)
library(ggplot2)
library(patchwork)
```


## Data 

```{r}
#| label: cleaning and wrangling
#| message: false
#| warning: false

compiled_isotope <- read_csv(here("data", "compiled.csv"), show_col_types = FALSE) |> 
  clean_names() |> # lowercase and underscores in column names
  rename(sample = sample_name, # renaming columns to be shorter 
         delta_d = processed_delta_d,
         delta_d_std = processed_delta_d_stdev,
         delta_o = processed_delta_18o,
         delta_o_std = processed_delta_18o_stdev) |> 
  mutate(date = mdy(date)) # telling r the date format

gradient_tc <- read_csv(here("data", "tc_gradient_data.csv"))

gradient_cc <- read_csv(here("data", "cc_gradient_data.csv"))

metadata <- read_csv(here("data", "metadata.csv"))

trail_creek <- compiled_isotope |> 
  filter(location %in% c("utc", "tc"))

coal_creek <- compiled_isotope |> 
  filter(location %in% c("cc")) |> 
  filter(site != "coal_3")

coal_creek_all <- compiled_isotope |> 
 filter(location %in% c("cc"))

italian_creek <- compiled_isotope |> 
  filter(location == "ic")


# Downstream Trail Creek 
tc_data <- read_csv(here("data", "TrailCreek_2026-07-23-15.csv"), show_col_types = FALSE) |> # reading in data
  clean_names() |> # lowercase and underscores in column names
  rename(date_time = date_time_mdt, # renaming columns to be shorter 
         dp = differential_pressure_k_pa,
         ap = absolute_pressure_k_pa,
         temp = temperature_c,
         water_level = water_level_m,
         bp = barometric_pressure_k_pa) |> 
  mutate(date_time_parsed = mdy_hms(date_time), # telling r the date format
         date = as_date(date_time_parsed)) |> # making a date column 
  group_by(date) |> 
  summarise(avg_dp = mean(dp, na.rm = TRUE), # averaging the data based on the day
            avg_ap = mean(ap, na.rm = TRUE),
            avg_temp = mean(temp, na.rm = TRUE),
            avg_water_level = mean(water_level, na.rm = TRUE),
            avg_bp = mean(bp, na.rm = TRUE),
            total_readings = n(), # the amount of data it used to average
            .groups = "drop") |>
  mutate(site = "tc") # creating a new column of the site 

# Upstream Trail Creek
utc_data <- read_csv(here("data", "UTC_2026-07-23-15.csv"), show_col_types = FALSE) |> # reading in data
  clean_names() |> # lowercase and underscores in column names
  rename(date_time = date_time_mdt,  # renaming columns to be shorter 
         dp = differential_pressure_k_pa,
         ap = absolute_pressure_k_pa,
         temp = temperature_c,
         water_level = water_level_m,
         bp = barometric_pressure_k_pa) |>
  mutate(date_time_parsed = mdy_hms(date_time), # telling r the date format
         date = as_date(date_time_parsed)) |> # making a date column 
  group_by(date) |> 
  summarise(avg_dp = mean(dp, na.rm = TRUE),  # averaging the data based on the day
            avg_ap = mean(ap, na.rm = TRUE),
            avg_temp = mean(temp, na.rm = TRUE),
            avg_water_level = mean(water_level, na.rm = TRUE),
            avg_bp = mean(bp, na.rm = TRUE),
            total_readings = n(), # the amount of data it used to average
            .groups = "drop") |>
  mutate(site = "utc") # creating a new column of the site 

tc_utc_data <- merge(tc_data, utc_data, all = TRUE) # merging the tc and utc data 

isco_data <- merge(
  x = tc_utc_data, 
  y = compiled_isotope, 
  by.x = c("site", "date"), # columns in tc_utc_data
  by.y = c("location", "date") # matching columns in compiled_isotope
)



tc_isco <- isco_data |> 
  filter(site %in% c("tc"))

utc_isco <- isco_data |> 
  filter(site %in% c("utc"))


```

## Statistical Tests 

```{r}
#| label: stat tests
#| message: false
#| warning: false

#----------------------- ANOVA of lc-excess and location ---------------------- #
anova <- aov(lc_excess ~ location, data = compiled_isotope)
summary(anova)

TukeyHSD(anova)

#----------------------- Correlation od location and date  ---------------------- #

compiled_isotope$date_c <- as.numeric(compiled_isotope$date) - mean(as.numeric(compiled_isotope$date))
# center the date variable by subtracting its mean.

model <- lmer(
  lc_excess ~ location + date_c + # fixed effects: location differences + linear time trend
    (1|site), # random intercept: allows each site to have its own baseline lc-excess, accounting for
              # the ~36% of variance that's due to site-level clustering rather than location or date
  data = compiled_isotope)

summary(model)
# centering doesn't affect slopes, only the intercept's ^ from the (1|site)

anova(model)


#----------------------- Correlation with normalized stream distance ---------------------- #

# Trail Creek 6/24 - lc-excess & cumulative dams
model_6_24_str_tc <- lm(`lc-excess_6-24` ~ normalized_stream_distance_m, data = gradient_tc)
summary(model_6_24_str_tc)

# Trail Creek 6/30 - lc-excess & cumulative dams
model_6_30_str_tc <-  lm(`lc-excess_6-30` ~ normalized_stream_distance_m, data = gradient_tc)
summary(model_6_30_str_tc)

# Trail Creek 7/15 - lc-excess & cumulative dams
model_7_15_str_tc <- lm(`lc-excess_7-15` ~ normalized_stream_distance_m, data = gradient_tc)
summary(model_7_15_str_tc)

#__________________________________________________________________________________#

# Coal Creek 6/23 - lc-excess & cumulative dams
model_6_23_str_cc <-  lm(`lc-excess_6-23` ~ normalized_stream_distance_m, data = gradient_cc)
summary(model_6_23_str_cc)

# # Coal Creek 7/7 - lc-excess & cumulative dams
model_7_7_str_cc <-  lm(`lc-excess_7-7` ~ normalized_stream_distance_m, data = gradient_cc)
summary(model_7_7_str_cc)

#----------------------- Correlation of Trail Creek Upstream and Downstream  ---------------------- #
cor.test(tc_isco$avg_water_level, tc_isco$lc_excess)

cor.test(utc_isco$avg_water_level, utc_isco$lc_excess)


#----------------------- Correlation with cumulative dams ---------------------- #
gradient_tc$cumulative_dams <- as.numeric(as.character(gradient_tc$cumulative_dams))

# Trail Creek 6/24 - lc-excess & cumulative dams
model_6_24_cum_tc <- lm(`lc-excess_6-24` ~ cumulative_dams, data = gradient_tc)
summary(model_6_24_cum_tc)

# Trail Creek 6/30 - lc-excess & cumulative dams
model_6_30_cum_tc <-  lm(`lc-excess_6-30` ~ cumulative_dams, data = gradient_tc)
summary(model_6_30_cum_tc)

# Trail Creek 7/15 - lc-excess & cumulative dams
model_7_15_cum_tc <- lm(`lc-excess_7-15` ~ cumulative_dams, data = gradient_tc)
summary(model_7_15_cum_tc)

#__________________________________________________________________________________#

gradient_cc$cumulative_dams <- as.numeric(as.character(gradient_cc$cumulative_dams))

# Coal Creek 6/23 - lc-excess & cumulative dams
model_6_23_cum_cc <-  lm(`lc-excess_6-23` ~ cumulative_dams, data = gradient_cc)
summary(model_6_23_cum_cc)

# # Coal Creek 7/7 - lc-excess & cumulative dams
model_7_7_cum_cc <-  lm(`lc-excess_7-7` ~ cumulative_dams, data = gradient_cc)
summary(model_7_7_cum_cc)

#----------------------- Correlation with cumulative dam length ---------------------- #
gradient_tc$cumulative_dam_length <- as.numeric(as.character(gradient_tc$cumulative_dam_length))

# Trail Creek 6/24 - lc-excess & cumulative dam length
model_6_24_cumd_tc <- lm(`lc-excess_6-24` ~ cumulative_dam_length, data = gradient_tc)
summary(model_6_24_cumd_tc)

# Trail Creek 6/30 - lc-excess & cumulative dam length
model_6_30_cumd_tc <-  lm(`lc-excess_6-30` ~ cumulative_dam_length, data = gradient_tc)
summary(model_6_30_cumd_tc)

# Trail Creek 7/15 - lc-excess & cumulative dam length
model_7_15_cumd_tc <- lm(`lc-excess_7-15` ~ cumulative_dam_length, data = gradient_tc)
summary(model_7_15_cumd_tc)

#__________________________________________________________________________________#

gradient_cc$cumulative_dam_length <- as.numeric(as.character(gradient_cc$cumulative_dam_length))

# Coal Creek 6/23 - lc-excess & cumulative dam length
model_6_23_cumd_cc <-  lm(`lc-excess_6-23` ~ cumulative_dam_length, data = gradient_cc)
summary(model_6_23_cumd_cc)

# Coal Creek 7/7 - lc-excess & cumulative dam length
model_7_7_cumd_cc <-  lm(`lc-excess_7-7` ~ cumulative_dam_length, data = gradient_cc)
summary(model_7_7_cumd_cc)

#----------------------- Correlation with average water level ---------------------- #


# Trail Creek - lc-excess & average water level
model_water_tc <- lm(lc_excess ~ avg_water_level, data = tc_isco) 
summary(model_water_tc)

# Upstream Trail Creek (has no beaver dams)- lc-excess & average water level
model_water_utc <- lm(lc_excess ~ avg_water_level, data = utc_isco)
summary(model_water_utc)



```


## Exploratory Visuals with ALL Sites

Line-Conditioned Excess (lc-excess) vs Delta 18O 

```{r}
#| message: false
#| warning: false
#| fig-height: 5
#| fig-width: 6

ggplot(data = compiled_isotope, aes(x = delta_o, y = lc_excess, color = location)) +
  geom_point(size = 2.5, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.15, linewidth = 1) + 
  scale_color_manual(
    values = c("tc" = "royalblue2", "cc" = "turquoise3", "it" = "deeppink2", "utc" = "orangered2"),
    labels = c("cc" = "Coal Creek", "it" = "Italian Creek", "tc" = "Trail Creek", "utc" = "Upper Trail Creek")
  ) +
  labs(
    title = "Line-Conditioned Excess (lc-excess) vs. Delta 18O",
    subtitle = "Comparing evaporation signals between sites",
    x = expression(paste(delta, " "^18, "O (\u2030)")), # Formats as δ18O (‰)
    y = "lc-excess (\u2030)",
    color = "Site Location") +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 14),
        legend.position = "bottom",
        panel.grid.minor = element_blank())

```

Time-Series of Line-Conditioned Excess (lc-excess)

```{r}
#| message: false
#| warning: false

ggplot(data = compiled_isotope, 
       aes(x = date, 
           y = lc_excess, 
           color = location, 
           group = location)) +
  geom_hline(yintercept = 0, 
             linetype = "dashed", 
             color = "gray50", 
             size = 0.8) +
  geom_point(size = 2.5, 
             alpha = 0.8) +
  geom_line(linewidth = 1, 
            alpha = 0.7) +
  labs(title = "Time-Series of Line-Conditioned Excess (lc-excess)",
       subtitle = "Values below 0 indicate potential evaporative enrichment",
       x = "Sampling Date",
       y = "lc-excess (\u2030)",
       color = "Site") +
  ylim(-8,5) +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 14),
        panel.grid.minor = element_blank(),
        legend.position = "top")

```

Time-Series of Deuterium Excess (d-excess)

```{r}
#| message: false
#| warning: false

ggplot(data = compiled_isotope, 
       aes(x = date, 
           y = d_excess, 
           color = location, 
           group = location)) +
  geom_hline(yintercept = 0, 
             linetype = "dashed", 
             color = "gray50", 
             size = 0.8) +
  geom_point(size = 2.5, 
             alpha = 0.8) +
  geom_line(size = 1, 
            alpha = 0.7) + 
  labs(title = "Time-Series of Deuterium Excess (d-excess)",
       subtitle = "Values below 0 indicate potential evaporative enrichment",
       x = "Sampling Date",
       y = "d-excess (\u2030)",
       color = "Site") +
  ylim(-3,15) +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 14),
        panel.grid.minor = element_blank(),
        legend.position = "top")

```

Temporal Variations in Line-Conditioned Excess (by location)

```{r}
#| label: line
#| message: false
#| warning: false

ggplot(compiled_isotope, aes(x = date, y = lc_excess, color = location)) +
  geom_point(size = 1.5) +
  labs(
    title = "Temporal Variations in Line-Conditioned Excess",
    x = "Date",
    y = "lc-excess (‰)",
    color = "Location"
  ) +
  facet_wrap(~location, scales = "free_y") +  # Compare across sites easily
  theme_minimal() + 
  theme(plot.title = element_text(face = "bold", size = 14))

```

Temporal Variations in Deuterium Excess (by sites)

```{r}
#| label: timeseries
#| message: false
#| warning: false

ggplot(compiled_isotope, aes(x = date, y = d_excess, color = location)) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1.5) +
  labs(
    title = "Temporal Variations in Deuterium Excess",
    x = "Date",
    y = "lc-excess (‰)",
    color = "Site"
  ) +
  facet_wrap(~site, scales = "free_y") +  # Compare across sites easily
  theme_minimal() + 
  theme(plot.title = element_text(face = "bold", size = 14))

```

Monthly Variations in Evaporative Fraction (lc-excess) Across Study Sites

```{r}
#| label: boxplot-monthly
#| message: false
#| warning: false

ggplot(compiled_isotope, aes(x = factor(month(date, label = TRUE, abbr = TRUE)), y = lc_excess)) +
  geom_boxplot(aes(fill = factor(month(date))), show.legend = FALSE) +
  facet_wrap(~location, labeller = labeller(location = c(tc = "Trail Creek", 
                                                         cc = "Coal Creek",
                                                         it = "Italian Creek",
                                                         utc = "Upstream Trail Creek"))) +
 labs(
    title = "Monthly Variations in Evaporative Fraction (lc-excess) Across Study Sites",
    subtitle = "Seasonal isotopic shifts indicating changes in local evaporation-condensation histories",
    x = "Month",
    y = "lc-excess (‰)"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 11),
    axis.text.x = element_text(angle = 45, hjust = 1)) # tilting the x axis text

```


Time series of lc-excess across season, colored by reach, faceted by watershed

```{r}
#| label: seasonal-trends-in-lcexcess
#| message: false
#| warning: false

ggplot(compiled_isotope, aes(x = date, y = lc_excess, color = site, group = interaction(site, location))) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(size = 2, alpha = 0.8) +
  geom_line(alpha = 0.6) +
  facet_wrap(~location, labeller = labeller(location = c(tc = "Trail Creek", 
                                                         cc = "Coal Creek",
                                                         it = "Italian Creek",
                                                         utc = "Upstream Trail Creek"))) +
  scale_x_date(date_labels = "%b %d") +
  scale_color_manual(values = wes_palette("FantasticFox1", n = length(unique(compiled_isotope$site)),
                                          type = "continuous"),
                     labels = c("coal_1" = "CC-1", "coal_2" = "CC-2", "coal_3" = "CC-3", "coal_4" =
                                "CC-4", "coal_5" = "CC-5", "coal_6" = "CC-6", "italian_1" = "IT-1",
                                "italian_2" = "IT-2", "italian_3" = "IT-3", "italian_4" = "IT-4",
                                "italian_5" = "IT-5","tc_isco" = "TC-ISCO", "tc_1" = "TC-1", "tc_15"
                                = "TC-1.5", "tc_2" = "TC-2", "tc_3" = "TC-3", "tc_ds" = "TC-DS",
                                "tc_mid" = "TC-Mid", "tc_us" = "TC-US", "utc_isco" = "UTC-ISCO")) +
  labs(title = "Seasonal Trends in lc-excess",
       subtitle = "2025 low-snowpack year",
       x = "Sampling Date", y = "lc-excess (\u2030)", color = "Reach") +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 14),
        legend.position = "right",
        axis.text.x = element_text(angle = 45, hjust = 1))


```

Dual-isotope plot (δ18O vs δD) with LMWL reference

```{r}
#| label: dual
#| message: false
#| warning: false

# Add Global Meteoric Water Line (GMWL: y = 8x + 10)

gmwl_slope <- 8.00
gmwl_intercept <- 10.00 

ggplot(compiled_isotope, aes(x = delta_o, y = delta_d, color = site)) +
  geom_point(size = 2.5, alpha = 0.7) +
  geom_abline(slope = gmwl_slope, intercept = gmwl_intercept, 
              linetype = "dashed", color = "black", linewidth = 0.8) +
  facet_wrap(~location, labeller = labeller(location = c(tc = "Trail Creek",
                                                         utc = "Upstream Trail Creek",
                                                         cc = "Coal Creek",
                                                         it = "Italian Creek"))) +
  labs(title = "Dual Isotope Plot: \u03b418O vs \u03b4D",
       subtitle = "Dashed line = Global Meteoric Water Line",
       x = expression(paste(delta^18, "O (\u2030)")),
       y = expression(paste(delta, "D (\u2030)")),
       color = "Reach") +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 14),
        legend.position = "none")

```


