---
title: "all-sites-exploratory"
author: "Linda Herrera"
format: typst
execute:
  fig-format: png
---


## Packages

```{r}
#| label: packages
#| message: false
#| warning: false

library(tidyverse)
library(here)
library(janitor)
library(viridis)
library(wesanderson)
```


## Data 

```{r}
#| label: cleaning and wrangling
#| message: false
#| warning: false

compiled_isotope <- read_csv(here("data", "compiled.csv"), show_col_types = FALSE) |> 
  clean_names() |> # lowercase and underscores in column names
  rename(sample = sample_name, # renaming columns to be shorter 
         delta_d = processed_delta_d,
         delta_d_std = processed_delta_d_stdev,
         delta_o = processed_delta_18o,
         delta_o_std = processed_delta_18o_stdev) |> 
  mutate(date = mdy(date)) # telling r the date format

gradient_tc <- read_csv(here("data", "tc_gradient_data.csv"))

gradient_cc <- read_csv(here("data", "cc_gradient_data.csv"))

metadata <- read_csv(here("data", "metadata.csv"))

trail_creek <- compiled_isotope |> 
  filter(location %in% c("utc", "tc"))

coal_creek <- compiled_isotope |> 
  filter(location %in% c("cc")) |> 
  filter(site != "coal_3")

coal_creek_all <- compiled_isotope |> 
 filter(location %in% c("cc"))

italian_creek <- compiled_isotope |> 
  filter(location == "ic")

```

## Exploratory Visuals with ALL Sites

Line-Conditioned Excess (lc-excess) vs Delta 18O 

```{r}
#| message: false
#| warning: false
#| fig-height: 5
#| fig-width: 6

ggplot(data = compiled_isotope, aes(x = delta_o, y = lc_excess, color = location)) +
  geom_point(size = 2.5, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.15, linewidth = 1) + 
  scale_color_manual(
    values = c("tc" = "royalblue2", "cc" = "turquoise3", "ic" = "deeppink2", "utc" = "orangered2"),
    labels = c("cc" = "Coal Creek", "ic" = "Italian Creek", "tc" = "Trail Creek", "utc" = "Upper Trail Creek")
  ) +
  labs(
    title = "Line-Conditioned Excess (lc-excess) vs. Delta 18O",
    subtitle = "Comparing evaporation signals between sites",
    x = expression(paste(delta, " "^18, "O (\u2030)")), # Formats as δ18O (‰)
    y = "lc-excess (\u2030)",
    color = "Site Location") +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 14),
        legend.position = "bottom",
        panel.grid.minor = element_blank())

```

Time-Series of Line-Conditioned Excess (lc-excess)

```{r}
#| message: false
#| warning: false

ggplot(data = compiled_isotope, 
       aes(x = date, 
           y = lc_excess, 
           color = location, 
           group = location)) +
  geom_hline(yintercept = 0, 
             linetype = "dashed", 
             color = "gray50", 
             size = 0.8) +
  geom_point(size = 2.5, 
             alpha = 0.8) +
  geom_line(linewidth = 1, 
            alpha = 0.7) +
  labs(title = "Time-Series of Line-Conditioned Excess (lc-excess)",
       subtitle = "Values below 0 indicate potential evaporative enrichment",
       x = "Sampling Date",
       y = "lc-excess (\u2030)",
       color = "Site") +
  ylim(-8,5) +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 14),
        panel.grid.minor = element_blank(),
        legend.position = "top")

```

Time-Series of Deuterium Excess (d-excess)

```{r}
#| message: false
#| warning: false

ggplot(data = compiled_isotope, 
       aes(x = date, 
           y = d_excess, 
           color = location, 
           group = location)) +
  geom_hline(yintercept = 0, 
             linetype = "dashed", 
             color = "gray50", 
             size = 0.8) +
  geom_point(size = 2.5, 
             alpha = 0.8) +
  geom_line(size = 1, 
            alpha = 0.7) + 
  labs(title = "Time-Series of Deuterium Excess (d-excess)",
       subtitle = "Values below 0 indicate potential evaporative enrichment",
       x = "Sampling Date",
       y = "d-excess (\u2030)",
       color = "Site") +
  ylim(-3,15) +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 14),
        panel.grid.minor = element_blank(),
        legend.position = "top")

```

Temporal Variations in Line-Conditioned Excess (by location)

```{r}
#| label: line
#| message: false
#| warning: false

ggplot(compiled_isotope, aes(x = date, y = lc_excess, color = location)) +
  geom_point(size = 1.5) +
  labs(
    title = "Temporal Variations in Line-Conditioned Excess",
    x = "Date",
    y = "lc-excess (‰)",
    color = "Location"
  ) +
  facet_wrap(~location, scales = "free_y") +  # Compare across sites easily
  theme_minimal() + 
  theme(plot.title = element_text(face = "bold", size = 14))

```

Temporal Variations in Deuterium Excess (by sites)

```{r}
#| label: timeseries
#| message: false
#| warning: false

ggplot(compiled_isotope, aes(x = date, y = d_excess, color = location)) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1.5) +
  labs(
    title = "Temporal Variations in Deuterium Excess",
    x = "Date",
    y = "lc-excess (‰)",
    color = "Site"
  ) +
  facet_wrap(~site, scales = "free_y") +  # Compare across sites easily
  theme_minimal() + 
  theme(plot.title = element_text(face = "bold", size = 14))

```

Monthly Variations in Evaporative Fraction (lc-excess) Across Study Sites

```{r}
#| label: boxplot-monthly
#| message: false
#| warning: false

ggplot(compiled_isotope, aes(x = factor(month(date, label = TRUE, abbr = TRUE)), y = lc_excess)) +
  geom_boxplot(aes(fill = factor(month(date))), show.legend = FALSE) +
  facet_wrap(~location, scales = "free_y") +
 labs(
    title = "Monthly Variations in Evaporative Fraction (lc-excess) Across Study Sites",
    subtitle = "Seasonal isotopic shifts indicating changes in local evaporation-condensation histories",
    x = "Month",
    y = "lc-excess (‰)"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 11),
    axis.text.x = element_text(angle = 45, hjust = 1)) # tilting the x axis text

```

