---
title: "all-sites-exploratory"
author: "Linda Herrera"
format: typst
execute:
  fig-format: png
---


## Packages

```{r}
#| label: packages
#| message: false
#| warning: false

library(tidyverse)
library(here)
library(janitor)
library(snakecase)
library(scales)
library(readxl)
library(patchwork)
library(ggridges)
library(lme4)
library(lmerTest)
library(knitr)
library(rmarkdown)
library(viridis)
library(viridisLite)
library(wesanderson)
library(ragg)

```


## Data 

```{r}
#| label: cleaning and wrangling
#| message: false
#| warning: false

# Trail Creek Isotope 
compiled_isotope <- read_csv(here("data", "compiled.csv"), show_col_types = FALSE) |> 
  clean_names() |> # lowercase and underscores in column names
  rename(sample = sample_name, # renaming columns to be shorter 
         delta_d = processed_delta_d,
         delta_d_std = processed_delta_d_stdev,
         delta_o = processed_delta_18o,
         delta_o_std = processed_delta_18o_stdev) |> 
  mutate(date = mdy(date)) # telling r the date format

gradient_tc <- read_csv(here("data", "tc_gradient_data.csv"))

gradient_cc <- read_csv(here("data", "cc_gradient_data.csv"))

metadata <- read_csv(here("data", "metadata.csv"))

trail_creek <- compiled_isotope |> 
  filter(location %in% c("utc", "tc"))

coal_creek <- compiled_isotope |> 
  filter(location %in% c("cc")) |> 
  filter(site != "coal_3")

coal_creek_all <- compiled_isotope |> 
  filter(location %out% c("cc"))

italian_creek <- compiled_isotope |> 
  filter(location == "ic")

```

## Exploratory Visuals with ALL Sites

Line-Conditioned Excess (lc-excess) vs Delta 18O 

```{r}
#| message: false
#| warning: false
#| fig-height: 5
#| fig-width: 6

ggplot(data = compiled_isotope, aes(x = delta_o, y = lc_excess, color = location)) +
  geom_point(size = 2.5, alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.15, linewidth = 1) + 
  scale_color_manual(
    values = c("tc" = "royalblue2", "cc" = "turquoise3", "ic" = "deeppink2", "utc" = "orangered2"),
    labels = c("cc" = "Coal Creek", "ic" = "Italian Creek", "tc" = "Trail Creek", "utc" = "Upper Trail Creek")
  ) +
  labs(
    title = "Line-Conditioned Excess (lc-excess) vs. Delta 18O",
    subtitle = "Comparing evaporation signals between sites",
    x = expression(paste(delta, " "^18, "O (\u2030)")), # Formats as δ18O (‰)
    y = "lc-excess (\u2030)",
    color = "Site Location") +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 14),
        legend.position = "bottom",
        panel.grid.minor = element_blank())

```

Time-Series of Line-Conditioned Excess (lc-excess)

```{r}
#| message: false
#| warning: false

ggplot(data = compiled_isotope, 
       aes(x = date, 
           y = lc_excess, 
           color = location, 
           group = location)) +
  geom_hline(yintercept = 0, 
             linetype = "dashed", 
             color = "gray50", 
             size = 0.8) +
  geom_point(size = 2.5, 
             alpha = 0.8) +
  geom_line(linewidth = 1, 
            alpha = 0.7) +
  labs(title = "Time-Series of Line-Conditioned Excess (lc-excess)",
       subtitle = "Values below 0 indicate potential evaporative enrichment",
       x = "Sampling Date",
       y = "lc-excess (\u2030)",
       color = "Site") +
  ylim(-8,5) +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 14),
        panel.grid.minor = element_blank(),
        legend.position = "top")

```

Time-Series of Deuterium Excess (d-excess)

```{r}
#| message: false
#| warning: false

ggplot(data = compiled_isotope, 
       aes(x = date, 
           y = d_excess, 
           color = location, 
           group = location)) +
  geom_hline(yintercept = 0, 
             linetype = "dashed", 
             color = "gray50", 
             size = 0.8) +
  geom_point(size = 2.5, 
             alpha = 0.8) +
  geom_line(size = 1, 
            alpha = 0.7) + 
  labs(title = "Time-Series of Deuterium Excess (d-excess)",
       subtitle = "Values below 0 indicate potential evaporative enrichment",
       x = "Sampling Date",
       y = "d-excess (\u2030)",
       color = "Site") +
  ylim(-3,15) +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 14),
        panel.grid.minor = element_blank(),
        legend.position = "top")

```

Temporal Variations in Line-Conditioned Excess (by location)

```{r}
#| label: line
#| message: false
#| warning: false

ggplot(compiled_isotope, aes(x = date, y = lc_excess, color = location)) +
  geom_point(size = 1.5) +
  labs(
    title = "Temporal Variations in Line-Conditioned Excess",
    x = "Date",
    y = "lc-excess (‰)",
    color = "Location"
  ) +
  facet_wrap(~location, scales = "free_y") +  # Compare across sites easily
  theme_minimal() + 
  theme(plot.title = element_text(face = "bold", size = 14))

```

Temporal Variations in Deuterium Excess (by sites)

```{r}
#| label: timeseries
#| message: false
#| warning: false

ggplot(compiled_isotope, aes(x = date, y = d_excess, color = location)) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1.5) +
  labs(
    title = "Temporal Variations in Deuterium Excess",
    x = "Date",
    y = "lc-excess (‰)",
    color = "Site"
  ) +
  facet_wrap(~site, scales = "free_y") +  # Compare across sites easily
  theme_minimal() + 
  theme(plot.title = element_text(face = "bold", size = 14))

```

Monthly Variations in Evaporative Fraction (lc-excess) Across Study Sites

```{r}
#| label: boxplot-monthly
#| message: false
#| warning: false

ggplot(compiled_isotope, aes(x = factor(month(date, label = TRUE, abbr = TRUE)), y = lc_excess)) +
  geom_boxplot(aes(fill = factor(month(date))), show.legend = FALSE) +
  facet_wrap(~location, scales = "free_y") +
 labs(
    title = "Monthly Variations in Evaporative Fraction (lc-excess) Across Study Sites",
    subtitle = "Seasonal isotopic shifts indicating changes in local evaporation-condensation histories",
    x = "Month",
    y = "lc-excess (‰)"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 11),
    axis.text.x = element_text(angle = 45, hjust = 1)) # tilting the x axis text

```

## Exploratory Visuals of Trail Creek

Line-Conditioned Excess (lc-excess) vs. Delta 18O

```{r}
#| label: scatterplot
#| message: false
#| warning: false

ggplot(data = trail_creek, aes(x = delta_o, y = lc_excess, color = site)) +
  geom_point(size = 2.5, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, alpha = 0.15, linewidth = 1) +
  labs(
    title = "Line-Conditioned Excess (lc-excess) vs. Delta 18O",
    subtitle = "Comparing evaporation signals between sites",
    x = expression(paste(delta, " "^18, "O (\u2030)")), # Formats as δ18O (‰)
    y = "lc-excess (\u2030)",
    color = "Site Location") +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 14),
        legend.position = "right",
        panel.grid.minor = element_blank())

```

Time-Series of Line-Conditioned Excess (lc-excess)

```{r}
#| label: time-series
#| message: false
#| warning: false

ggplot(data = trail_creek, 
       aes(x = date, 
           y = lc_excess)) +
  geom_hline(yintercept = 0, 
             linetype = "dashed", 
             color = "gray50", 
             size = 0.8) +
  geom_point(size = 2.5, 
             alpha = 0.8) +
  geom_line(size = 1, 
            alpha = 0.7) +
  labs(title = "Time-Series of Line-Conditioned Excess (lc-excess)",
       subtitle = "Values below 0 indicate potential evaporative enrichment",
       x = "Sampling Date",
       y = "lc-excess (\u2030)",
       color = "Site") +
  ylim(-8,5) +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 14),
        panel.grid.minor = element_blank(),
        legend.position = "top")

```

Time-Series of Deuterium Excess (d-excess)

```{r}
#| label: dexcess
#| message: false
#| warning: false

ggplot(data = trail_creek, 
       aes(x = date, 
           y = d_excess)) +
  geom_hline(yintercept = 0, 
             linetype = "dashed", 
             color = "gray50", 
             size = 0.8) +
  geom_point(size = 2.5, 
             alpha = 0.8) +
  geom_line(size = 1, 
            alpha = 0.7) + 
  labs(title = "Time-Series of Deuterium Excess (d-excess)",
       subtitle = "Values below 0 indicate potential evaporative enrichment",
       x = "Sampling Date",
       y = "d-excess (\u2030)",
       color = "Site") +
  ylim(-3,15) +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 14),
        panel.grid.minor = element_blank(),
        legend.position = "top")

```

Temporal Variations in Line-Conditioned Excess

```{r}
#| label: abc
#| message: false
#| warning: false

ggplot(trail_creek, aes(x = date, y = lc_excess, color = site)) +
  geom_line(linewidth = 0.5) +
  geom_point(size = 1.5) +
  labs(
    title = "Temporal Variations in Line-Conditioned Excess",
    x = "Date",
    y = "lc-excess (‰)",
    color = "Site"
  ) +
  facet_wrap(~site, scales = "free_y") +  # Compare across sites easily
  theme_minimal() + 
  theme(plot.title = element_text(face = "bold", size = 14))

```

Monthly Variations in Evaporative Fraction (lc-excess) Across Study Sites

```{r}
#| label: -monthly
#| message: false
#| warning: false

ggplot(trail_creek, aes(x = factor(month(date, label = TRUE, abbr = TRUE)), y = lc_excess)) +
  geom_boxplot(aes(fill = factor(month(date))), show.legend = FALSE) +
  facet_wrap(~location, scales = "free_y") +
 labs(
    title = "Monthly Variations in Evaporative Fraction (lc-excess) Across Study Sites",
    subtitle = "Seasonal isotopic shifts indicating changes in local evaporation-condensation histories",
    x = "Month",
    y = "lc-excess (‰)"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14),
    strip.text = element_text(face = "bold", size = 11), # Emphasize site names
    axis.text.x = element_text(angle = 45, hjust = 1)   # Tilt month labels if they overlap
  )

```

Dual-isotope plot (δ18O vs δD) with LMWL reference

```{r}
#| label: dual
#| message: false
#| warning: false

# Trail Creek LMWL calculated from OPIC

lmwl_slope <- 7.94
lmwl_intercept <- 9.37 

ggplot(compiled_isotope, aes(x = delta_o, y = delta_d, color = site)) +
  geom_point(size = 2.5, alpha = 0.7) +
  geom_abline(slope = lmwl_slope, intercept = lmwl_intercept, 
              linetype = "dashed", color = "black", linewidth = 0.8) +
  facet_wrap(~location, labeller = labeller(location = c(tc = "Trail Creek", 
                                                           cc = "Coal Creek", 
                                                           ic = "Italian Creek"))) +
  labs(title = "Dual Isotope Plot: \u03b418O vs \u03b4D",
       subtitle = "Dashed line = Trail Creek LMWL",
       x = expression(paste(delta^18, "O (\u2030)")),
       y = expression(paste(delta, "D (\u2030)")),
       color = "Reach") +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 14),
        legend.position = "top")

```

Time series of lc-excess across season, colored by reach, faceted by watershed

```{r}
#| label: seasonal-trends-in-lcexcess
#| message: false
#| warning: false

ggplot(compiled_isotope, aes(x = date, y = lc_excess, color = site, group = interaction(site, location))) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(size = 2, alpha = 0.8) +
  geom_line(alpha = 0.6) +
  facet_wrap(~location, labeller = labeller(location = c(tc = "Trail Creek", 
                                                         cc = "Coal Creek",
                                                         ic = "Italian Creek"))) +
  scale_x_date(date_labels = "%b %d") +
  scale_color_manual(values = wes_palette("FantasticFox1", n = length(unique(compiled_isotope$site)),
                                          type = "continuous"),
                     labels = c("coal_1" = "CC-1", "coal_2" = "CC-2", "coal_3" = "CC-3", "coal_4" =
                                "CC-4", "coal_5" = "CC-5", "coal_6" = "CC-6", "italian_1" = "IT-1",
                                "italian_2" = "IT-2", "italian_3" = "IT-3", "italian_4" = "IT-4",
                                "italian_5" = "IT-5","tc_isco" = "TC-ISCO", "tc_1" = "TC-1", "tc_15"
                                = "TC-1.5", "tc_2" = "TC-2", "tc_3" = "TC-3", "tc_ds" = "TC-DS",
                                "tc_mid" = "TC-Mid", "tc_us" = "TC-US", "utc_isco" = "UTC-ISCO")) +
  labs(title = "Seasonal Trends in lc-excess",
       subtitle = "2025 low-snowpack year",
       x = "Sampling Date", y = "lc-excess (\u2030)", color = "Reach") +
  theme_bw() +
  theme(plot.title = element_text(face = "bold", size = 14),
        legend.position = "right",
        axis.text.x = element_text(angle = 45, hjust = 1))


```

```{r}

# averaging the lc-excess and including standard error
delta_summary <- compiled_isotope |> 
  group_by(site) |> 
  summarize(mean_lc = mean(lc_excess, na.rm = TRUE),
            se_lc = sd(lc_excess, na.rm = TRUE) / sqrt(n()),
            .groups = "drop")

# creating a plot to show the mean lc-excess of each site
ggplot(delta_summary, aes(x = site, y = mean_lc, fill = site)) +
  geom_col(color = "black", width = 0.7) +
  geom_errorbar(aes(ymin = mean_lc - se_lc, ymax = mean_lc + se_lc),
                width = 0.3,
                color = "black") +
  scale_x_discrete(limits = c("italian_1", "italian_2", "italian_3", "italian_4", "italian_5", "coal_1", "coal_2", "coal_3", "coal_4", "coal_5", "coal_6", "tc_isco", "tc_1", "tc_15", "tc_2", "tc_3", "utc_isco", "tc_ds", "tc_mid", "tc_us"),
  labels = c("coal_1" = "CC-1", "coal_2" = "CC-2", "coal_3" = "CC-3", "coal_4" = "CC-4",
             "coal_5" = "CC-5", "coal_6" = "CC-6", "italian_1" = "IT-1", "italian_2" = "IT-2",
             "italian_3" = "IT-3", "italian_4" = "IT-4", "italian_5" = "IT-5",
             "tc_isco" = "TC-ISCO", "tc_1" = "TC-1", "tc_15" = "TC-1.5", "tc_2" = "TC-2",
             "tc_3" = "IT-3", "tc_ds" = "IT-DS", "tc_mid" = "TC-Mid", "tc_us" = "TC-US",
             "utc_isco" = "UTC-ISCO")) +
  scale_fill_viridis_d() + # color palette 
  theme_minimal() +
  labs(title = "Mean LC Excess by Site with Standard Error",
       x = "Sampling Site",
       y = "Mean LC Excess") +
  theme(plot.title = element_text(face = "bold", size = 14),
        axis.text.x = element_text(angle = 45, hjust = 1), # tilt labels 
        legend.position = "none")



```



```{r}
#| label: gradient
#| message: false
#| warning: false
#| fig-height: 5
#| fig-width: 6

# Buildings a look up table of stream distances
distances <- tibble(site = c("TC-P3-US", "TC-P3-mid", "TC-P3-DS", "TC-BDA1",
                             "TC-BDA1.5", "TC-BDA2", "TC-BDA3"),
                    dist_m = c(57.6, 333, 654, 1160, 1620, 2240, 3150)) |> 
  mutate(xmax = dist_m,
         xmin = lag(dist_m, default = 0)) # previous site's distance, 0 for first

# Creating a filtered data frame for lc-excess
gradient_lc_clean <- gradient_data |>
  select(site, normalized_stream_distance_m, `Δlc-excess/Δx_6-24`, `Δlc-excess/Δx_6-30`) |> 
  pivot_longer( cols = c(`Δlc-excess/Δx_6-24`, `Δlc-excess/Δx_6-30`),
                names_to = "date",
                values_to = "gradient") |>
  mutate(date = case_when(date == "Δlc-excess/Δx_6-24" ~ "2026-06-24",
                          date == "Δlc-excess/Δx_6-30" ~ "2026-06-30"),
         gradient = as.numeric(gradient)) |>
  left_join(distances %>% select(site, xmin, xmax), by = "site") |> 
  mutate(site = factor(site, levels = distances$site)) |>
  filter(!is.na(gradient))

# Gradient plot
ggplot(gradient_lc_clean) +
  geom_rect(aes(xmin = xmin, xmax = xmax, ymin = 0, ymax = gradient),
            fill = "royalblue2", color = "black", linewidth = 0.3) +
  geom_hline(yintercept = 0, color = "royalblue2", linewidth = 0.4) +
  facet_wrap(~ date, nrow = 1) +
  scale_x_continuous(
    breaks = distances$dist_m,
    labels = distances$site) +
  labs(title = "Trail Creek: Spatial variations in Δlc-excess/Δx across sampling dates",
       x = "Normalized stream distance (m)",
       y = expression(paste(Delta, "lc-excess / ", Delta, "x (ppb/km)"))) +
  theme_bw(base_size = 11) +
  theme(plot.title = element_text(face = "bold", size = 14),
        panel.grid.minor = element_blank(),
        strip.text = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))


```

All reaches show lc-excess decreasing downstream, on both dates. Since lc-excess becoming more negative = more evaporative enrichment, this version says every single reach along Trail Creek is adding evaporative signal, with no recovery zones. That's a materially different story than "some reaches enrich, some reaches dilute" — it now reads as continuous, cumulative evaporative loss along the whole restored gradient.

Where the deepest enrichment is happening differs by date:

6-24: TC-P3-DS reach is the sharpest dip (~-2.2), essentially a local hotspot just downstream of the mid-P3 restoration point.
6-30: TC-BDA1.5→BDA2 is now the sharpest dip (~-2.15), while TC-P3-DS has relaxed to a much smaller magnitude (~-0.55) compared to 6-24.

TC-BDA3 is the outlier reach on both dates — it's close to flat (near-zero gradient), meaning that final reach adds almost no additional evaporative signal, regardless of date. That's consistent between both panels and is probably your most robust finding here: whatever's happening upstream, the lowest reach isn't contributing much additional fractionation.


```{r}

gradient_dams_clean <- gradient_data %>%
  select(site, cumulative_dams,
         `Δlc-excess/Δx_6-24`, `Δlc-excess/Δx_6-30`) %>%
  pivot_longer(
    cols = c(`Δlc-excess/Δx_6-24`, `Δlc-excess/Δx_6-30`),
    names_to = "date",
    values_to = "gradient"
  ) %>%
  mutate(
    date = case_when(
      date == "Δlc-excess/Δx_6-24" ~ "2026-06-24",
      date == "Δlc-excess/Δx_6-30" ~ "2026-06-30"
    ),
    gradient = as.numeric(gradient),
    cumulative_dams = as.numeric(cumulative_dams),
    site = factor(site, levels = c("TC-P3-US", "TC-P3-mid", "TC-P3-DS",
                                    "TC-BDA1", "TC-BDA1.5", "TC-BDA2", "TC-BDA3"))
  ) %>%
  filter(!is.na(gradient))


ggplot(gradient_dams_clean, aes(x = site, y = gradient, fill = site)) +
  geom_col(color = "black", width = 0.7) +
  geom_hline(yintercept = 0, color = "steelblue", linewidth = 0.4) +
  facet_wrap(~ date, nrow = 1) +
  scale_fill_manual(values = wes_palette("FantasticFox1", n = 7, type = "continuous")) +
  scale_x_discrete(labels = function(x) {
    lookup <- setNames(gradient_dams_clean$cumulative_dams, gradient_dams_clean$site)
    lookup[x]
  }) +
  labs(
    x = "Cumulative dam count",
    y = expression(Delta*"lc-excess/"*Delta*"x"),
    fill = "Site"
  ) +
  theme_bw(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold", size = 11),
    strip.background = element_rect(fill = "gray85", color = NA),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    legend.title = element_text(face = "bold")
  )


```

## Exploratory Visuals of Coal Creek







## Exploratory Visuals of Italian Creek




